---
title: Don't Take it Personally
subtitle: Hypothesis Tests and Confidence Intervals
format:
  html:
    self_contained: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```
<a href="../Downloads/Application_Activities/AA_Hypothesis_Conf_Int.qmd" download="../Downloads/Application_Activities/AA_Hypothesis_Conf_Int.qmd" type="text/markdown" download class="button-download">
    Download QMD
</a>


# Introduction

In this activity, you will execute statistical hypothesis tests and generate confidence intervals for each of the Big 5 personality traits using data collected from a random sample of Brother Cannon's Math 221 students.  

Begin by describing the population of this analysis.  This will determine how broad our conclusions can be. Also include a table of the summary statistics and any visualizations that might tell a story about the data.  

Then, for each personality trait, include:

1. A statement of the null and alternative hypotheses and why you chose the alternative you did.
1. Choose alpha $\alpha = $
1. Check that your test requirements are satisfied (`n > 30` or `qqPlot(respons_variable)`)
1. Run the one-sample t-test and state your conclusion
1. Calculate a $1-\alpha$ level confidence interval and describe in words what it means

Recall that we can use "favstats()" to get summary statistics, boxplot() and histogram() to get visualizations, and the t.test() function to get hypothesis tests and confidence intervals.  **Be sure to label your plots' axes and include a title.**  

```{r}

# Load Libraries

library(tidyverse)
library(mosaic)
library(rio)
library(car)


# Load Data
big5 <- import('https://raw.githubusercontent.com/byuistats/Math221D_Cannon/master/Data/All_class_combined_personality_data.csv')

```


# Extroversion

[Write an intro paragraph that includes the null and alternative hypothesis]



Ho:  Mu = 50
Ha:  Mu > 50

alpha = .025

```{r}
# Extroversion

big5 %>%
  group_by(`Course Section`) %>%
  summarise(
    `Mean of Extro.` = mean(Extroversion, na.rm = TRUE), 
    `SD Extro.` = sd(Extroversion, na.rm = TRUE), 
    `Mean Consc.` = mean(Conscientiousness, na.rm=TRUE),
    `SD Consc.` = sd(Conscientiousness),
    Count = n()
  ) %>% knitr::kable()


### I'm making a single variable called "extrov" that drops the missing values.  You can do the same thing with the other traits to make analysis a little easier
extrov <- na.omit(big5$Extroversion) 

histogram(extrov)

# For Hypothesis Test:
t.test(extrov, mu = 50, alternative = "greater")


# For a Confidence Interval:
t.test(extrov, mu = 50, alternative = "two.sided", conf.level = 1-.025)$conf.int


```


## Conclusion

[Write a concluding paragraph that states your conclusion about the hypothesis test and includes an interpretation of the confidence interval in context of the research question.]

# Agreeableness

[Write an intro paragraph that includes the null and alternative hypothesis]


```{r}
# Agreeableness

```

## Conclusion

[Write a concluding paragraph that states your conclusion about the hypothesis test and includes an interpretation of the confidence interval in context of the research question.]

# Openness

[Write an intro paragraph that includes the null and alternative hypothesis]


```{r}
# Openness

```

## Conclusion

[Write a concluding paragraph that states your conclusion about the hypothesis test and includes an interpretation of the confidence interval in context of the research question.]

# Neuroticism

[Write an intro paragraph that includes the null and alternative hypothesis]


```{r}
# Neuroticism

```

## Conclusion

[Write a concluding paragraph that states your conclusion about the hypothesis test and includes an interpretation of the confidence interval in context of the research question.]

# Conscientiousness

[Write an intro paragraph that includes the null and alternative hypothesis]


```{r}
# Conscientiousness

```

## Conclusion

[Write a concluding paragraph that states your conclusion about the hypothesis test and includes an interpretation of the confidence interval in context of the research question.]

