---
title: "Data Wrangling - Application Activity"
subtitle: "Personality Revisited"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# Introduction

The Big 5 personality test is the most widely accepted tool for modelling personality in academic psychology.  It is based on decades of statistical analysis of personality descriptions across languages and cultures.  The [big 5 traits](https://en.wikipedia.org/wiki/Big_Five_personality_traits)  are:

1. Openness
2. Conscientiousness
3. Extroversion
4. Agreeableness
5. Neuroticism

Brother Cannon collected personality data on students for the past several semesters, including a few metrics that may be associated with personality traits.  

__NOTE__:  Scores for personality traits are given in *percentiles* relative to the general population.  

In this activity, you will practice the process for approaching a dataset outlined in [class](https://byuistats.github.io/Math221D_Course/2-Tidy_Data/05-Exploring_New_Data_with_Tidyverse.html):

  1. Load the data and libraries
  2. Explore the data and generate hypotheses
  3. Prepare the data for analysis
  4. Perform the appropriate analysis
  
Data preparation will include using the `filter()` function.  For now, analysis means creating good visualizations that tell a story using `ggplot()` and base R.  

# 1 - Load the Data and Libraries

Run the following code to load the necessary R libraries and import the big 5 personality data:  

```{r}
library(tidyverse)
library(car)
library(mosaic)
library(rio)

big5 <- import("https://github.com/byuistats/Math221D_Course/raw/main/Data/All_class_combined_personality_data.csv") %>% tibble()

```


# 2 - Explore the data and generate hypotheses

In this section, you will explore the dataset including drilling down into individual columns.  

Recall a few useful functions for data exploration:

1. `dim()` to get the number of rows and columns in a dataset
2. `table()` for getting counts of categorical data
3. `unique()` for getting a list of each of the distinct values of categorical data
4. `favstats()` to get summary statistics for quantitative data
5. `histogram()` to visualize the distribution of a single quantitative variable
6. `boxplot()` and/or `ggplot()` to compare the distributions of a quantitative variable for different groups

__NOTE__:  Unless specified, you may use base R functions OR GGPlot.  

```{r}

```


__QUESTION__:  How many students have responded to this survey? (i.e. how many rows)  
__ANSWER__:  

__QUESTION__:  What is the maximum number of languages spoken among respondents?  
__ANSWER__:  

__QUESTION__:  Create a histogram of Extroversion. Be sure to label your graph with a title and improved x and y axis labels.   
```{r}
# Put your histogram code here:  

```


__QUESTION__:  What is the highest score for Extroversion?      
__ANSWER__:  

__QUESTION__:  What problems, if any, do you notice?  
__ANSWER__:  

__QUESTION__:  Make a table of birth months:
```{r}
# Answer

```

__QUESTION__:  What problems, if any, do you notice?  
__ANSWER__:  



# 3 & 4 - Prepare data for analysis and Create Visualization

In this section, you will be asked to create a clean dataset for each visualization.  

Recall a few useful functions for data wrangling:

1. `filter()` to include or exclude specific rows
2. `select()` to include specific columns

## Extroversion

__QUESTION__:  Create a new dataset called `extro` that includes only columns for birth month and extroversion scores.  Make sure it only has values that are real.  

HINT: Extroversion is measured in percentiles, and you should already know what the months of the year are.

```{r}

extro <- 

```

__QUESTION__:  **USE GGPLOT** to create a side-by-side boxplot of Extroversion scores for all birth months.

```{r}

ggplot()

```

__QUESTION__:  Based on the boxplot, which month appears to be the least extroverted?  Explain your reasoning.  
__ANSWER__:  

__QUESTION__:    Based on the boxplot, which month appears to be the most extroverted?  Explain your reasoning.  
__ANSWER__:  


## Neuroticism

__QUESTION__:  Create a dataset called `neuro` that includes only the Section and Neuroticism columns:  

```{r}

neuro <- 
  
```

__QUESTION__:  **USE GGPLOT** to create a side-by-side boxplot comparing Neuroticism for all the different sections:

```{r}


```

__QUESTION__:    Based on the boxplot, which section appears to be the lowest in trait neuroticism?  Explain your reasoning.  
__ANSWER__:  

