---
title: "Introducing the Bivariate Data"
---
```{r setup, include=FALSE}
# Ignore this bit of code.  This makes it so that you can render the qmd file even when there are errors in your code.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```
<a href="https://raw.githubusercontent.com/byuistats/Math221D_Course/main/Student_Work/Workbooks/Bivariate_Data_Intro.qmd" download="Bivariate_Data_Intro.qmd" type="text/markdown" download class="button-download">
    Download QMD
</a>


```{r}

library(tidyverse)
library(mosaic)
library(rio)

geyser <- import('https://byuistats.github.io/BYUI_M221_Book/Data/OldFaithful.xlsx')
names(geyser)

math <- import('https://byuistats.github.io/BYUI_M221_Book/Data/MathSelfEfficacy.xlsx')
names(math)

```


# Correlation Coefficients

What does the correlation coefficient, r, quantify?

What does it mean if the correlation coefficient, r, is negative?


# Geyser Data

## Scatter plot

Make a scatter plot showing the relationship between wait time and the duration of the next eruption.

Which variable is the Explanatory variable?  Which is the Response?

```{r}

plot(Duration ~ Wait, data = geyser)

```

Before looking at the Correlation Coefficient, r, describe in words the direction and strength of the relationship.  Does it look linear?

## Correlation Coefficient

```{r}

cor(Duration ~ Wait, data = geyser)

```



# Geyser Data

## Scatter plot

Make a scatter plot showing the relationship between the self reported confidence rating and their test score.

Which variable is the Explanatory variable?  Which is the Response?


```{r}

plot(Score ~ ConfidenceRatingMean, data = math)

```

Before looking at the Correlation Coefficient, r, describe in words the direction and strength of the relationship.  Does it look linear?

## Correlation Coefficient

```{r}

cor(Score ~ ConfidenceRatingMean, data = math)

```







