---
title: Introducing R
subtitle: Summarizing Quantitative Data
format:
  html:
    self_contained: yes
---
```{r setup, include=FALSE}
# Ignore this bit of code.  This makes it so that you can render the qmd file even when there are errors in your code.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# Load the libraries
```{r, echo = FALSE, eval=FALSE}

library(rio)
library(mosaic)
library(tidyverse)
library(car)

```

# Read in Data

```{r}

old_faithful <- rio::import("https://byuistats.github.io/M221R/Data/old_faithful.xlsx")

gratitude <- rio::import("https://byuistats.github.io/M221R/Data/gratitude.xlsx")


```


# Old Faithful

## Calculate the Summary Statistics for Duration

```{r}


```

What is the mean duration time of Old Faithful eruptions?

What is the standard deviation of duration?


## Create a Historgram for Duration

Create a histogram and describe the shape of the distribution of duration:

```{r}


```



## Calculate Summary Statistics for Wait time

```{r}



```

What is the mean wait time between eruptions?

What is the maximum wait time between eruptions?

The middle 50% of wait times will be between what 2 numbers?


# Gratitude

Calculate the summary statistics for both groups:

```{r}


```

Which treatment group has the highest mean?

Which treatment group has the largest spread of happiness scores?

Create a graph with boxplots for each group:

```{r}


```

Which group has the lowest first quartile?



